---
title: "Effects of Race and Education on Voting Behavior"
author: "Shuyang Qiu"
thanks: "Code and data are available at: https://github.com/shuyangqiu/election_support."
date: today
date-format: long
abstract: "Data from the 2020 Cooperative Election Study was used to create a model that predicts whether a person voted for Trump or Biden in the 2020 Election based on their
race and education. The analysis suggests higher levels of education and being part of certain racial groups, namely Black and Asian, are correlated with a greater likelihood of voting for Biden. This research can allow policymakers better cater to their base and help individuals to better understand the relationship between socioeconomic factors and politics."
format: 
  pdf: 
    fig-pos: 'H'
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#### Workspace setup ####
library(tidyverse)
library(arrow)
library(here)
library(knitr)
library(modelsummary)
library(collapse)
library(marginaleffects)

# Read data
ces2020 <- read_parquet(here("outputs/data/cleaned_data.parquet"))

# Read model
political_preferences <- readRDS(here("outputs/models/political_preferences.rds"))
```


# Introduction
The United States has influence in every corner of the world, through trade, alliances, conflicts, or other means. This influence shifts over time as politicians change. One of the largest shifts occur when a new president is elected.

This paper analyzes post-election survey data to construct a model that attempts to predict the likelyhood of an individual to vote for Biden in the 2020 presidential election based on their level of education and their race.

The model suggests higher levels of education and being Black or Asian are positively correlated with the likelyhood of voting for Biden while low or no education
and being White is correlated with voting for Trump.

The data section (@sec-data) explains the dataset used, sampling methodology, how it was cleaned and analyzed, and visualizations of the data. 
The model section (@sec-model) explains the model used to fit the dataset.
The results section (@sec-results) explains the results of the analysis with visualizations.

# Data {#sec-data}
The data used in this paper is from the 2020 Cooperative Election Study (CES) [@DVN/E9N6PH_2021] and was retrieved through the `dataverse` [@dataverse] library.
The CES is an American annual stratified sample survey that includes questions about demographic information, voting behavior, and other aspects of American politics.
The 2020 CES survey was conducted online by YouGov with 61000 respondents. YouGov's matched random sampling methodology is further detailed in @CESGuide [pp. 13].

The data was then cleaned and analyzed using statistical programming language R [@r] and the `tidyverse` [@tidyverse], `janitor` [@janitor], `arrow` [@arrow], `stanarm` [@rstanarm], `knitr` [@knitr], `modelsummary` [@modelsummary], `collapse` [@rcollapse], `marginaleffects` [@marginaleffects], and `here` [@here] libraries.

Filtering only respondents who are registered to vote and voted for either Trump or Biden, we are left with 43554 respondents. @tbl-races and @tbl-education shows demographic information about the remaining respondents.


```{r}
#| label: tbl-races
#| tbl-cap: The distribution of races
#| echo: false

freq <- table(ces2020$race)
percent <- signif(prop.table(freq), 3) * 100
kable(cbind(freq, percent), col.names = c("Race", "Number of respondents", "Percentage"))
```
```{r}
#| label: tbl-education
#| tbl-cap: The distribution of education levels
#| echo: false

freq <- table(ces2020$education)
percent <- signif(prop.table(freq), 3) * 100
kable(cbind(freq, percent), col.names = c("Highest education", "Number of respondents", "Percentage"))
```


@tbl-vote shows the presidential candidates that the respondents voted for.

```{r}
#| label: tbl-vote
#| tbl-cap: The distribution of votes
#| echo: false

freq <- table(ces2020$voted_for)
percent <- signif(prop.table(freq), 3) * 100
kable(cbind(freq, percent), col.names = c("Voted for", "Number of respondents", "Percentage"))
```

@fig-ces shows presidential candidates that the respondents voted for by race and education level.
```{r}
#| label: fig-ces
#| fig-cap: The distribution of presidential preferences, by race, and highest education
#| echo: false

ces2020 |>
  ggplot(aes(x = education, fill = voted_for)) +
  stat_count(position = "dodge") +
  facet_wrap(facets = vars(race), scales = "free_x") +
  theme_minimal() +
  labs(
    x = "Highest education",
    y = "Number of respondents",
    fill = "Voted for"
  ) +
  coord_flip() +
  scale_fill_brewer(palette = "Set1") +
  theme(legend.position = "bottom")
```

# Model {#sec-model}

The linear regression model used was 
$$
\begin{aligned}
    y_i|\pi_i &\sim \operatorname{Bern}(\pi_i) \\
    \operatorname{logit}(\pi_i) & = \beta_0+\beta_1 \times \operatorname{race}_i + \beta_2 \times \operatorname{education}_i \\
    \beta_0 &\sim \operatorname{Normal}(0, 2.5)\\
    \beta_1 &\sim \operatorname{Normal}(0, 2.5)\\
    \beta_2 &\sim \operatorname{Normal}(0, 2.5)
\end{aligned}
$$

where $y_i = 0$ if the respondent prefers Trump and $y_i = 1$ if the respondent prefers Biden, $\operatorname{race}_i$ is the race of the respondent, and $\operatorname{education}_i$ is the education of the respondent.

A logistic regression was used since a respondent's presidential preference is binary. By inspection of @fig-ces, we expect correlations between race and presidential preference as well as a correlation between education and preference.

# Results {#sec-results}

The model coefficients (@tbl-model) suggests race has a considerable effect on presidential preference. Additionally, more educated people are likely to vote for Biden. @fig-model-confidence shows the credibility intervals of these predictors.
```{r}
#| label: tbl-model
#| tbl-cap: Whether a respondent is likely to vote for Biden based on their race and education
#| echo: false
#| warning: false

modelsummary(
  list("Support Biden" = political_preferences),
  statistic = "mad",
  )
```

@fig-vote-education shows the predicted probability of each participant voting for Biden and their actual vote by education. We see a positive correlation between education level and probability of voting for Biden, suggesting more educated people are more likely to vote for Biden. 
```{r}
#| label: fig-vote-education
#| fig-cap: Estimated probabilities of voting for Biden by education
#| echo: false

predictions <- 
  predictions(political_preferences) |>
  as_tibble()

predictions |>
  mutate(voted_for = factor(voted_for)) |>
  ggplot(aes(x = education, y = estimate, color = voted_for)) +
  ylim(0, 1) +
  geom_jitter(width = 0.5, height = 0.01, alpha = 0.3) +
  labs(
    x = "Education level",
    y = "Estimated probability of voting for Biden",
    color = "Actually voted for"
  ) +
  theme_classic() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom")
```

@fig-vote-race shows the predicted probability of each participant voting for Biden and their actual vote by race. We see differences in distribution for each race, suggesting race has a impact on the voting preferences of people. 
In particular, White people are more likely to vote for Trump while Black and Asian people are more likely to vote for Biden. Results about hispanics and other racial groups
are less conclusive.
```{r}
#| label: fig-vote-race
#| fig-cap: Estimated probabilities of voting for Biden by race
#| echo: false

predictions <- 
  predictions(political_preferences) |>
  as_tibble()

predictions |>
  mutate(voted_for = factor(voted_for)) |>
  ggplot(aes(x = race, y = estimate, color = voted_for)) +
  ylim(0, 1) +
  geom_jitter(width = 0.5, height = 0.01, alpha = 0.3) +
  labs(
    x = "Race",
    y = "Estimated probability of voting for Biden",
    color = "Actually voted for"
  ) +
  theme_classic() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom")
```

# Appendix
```{r}
#| label: fig-model-confidence
#| fig-cap: Credible intervals for predictors of support for Biden
#| echo: false

modelplot(political_preferences, conf_level = 0.9) +
  labs(x = "90 per cent credibility interval")
```

\newpage
# References


